A Quantitative Genetic Linked Selection Model of the Negative Selection Process in Humans

### 2023-07-19 manuscript/manuscript.tex
ms updates

The discrepancy between our theoretic B' maps and chromosome-wide simulations
that leads to the 5\% error occurs entirely around the drift-barrier domain of
$2Ns \approx 1$ (Figure \ref{fig:figure-1}D). The higher error in this domain
occurs across all mutation rates. We hypothesized this error may be due to
selective interference between segments that is not taken into account when we
numerically solve Equations \eqref{eq:main_eqns} for each segment
independently. In particular, we use a fixed drift-effective $N=1000$
corresponding to the number of diploids in the simulations rather than $B(x) N$
at position $x$. To test this, we implemented a locally-rescaled version of the
B' maps, which numerically solves Equations \eqref{eq:main_eqns} for each
segment at position $x$ taking into account the reduction experienced due to
selection at other segments by setting $B'(x)N$ as the population size. We find
the locally-rescaled B' maps reduce the relative error from 5\% to 0.4\% and
mean squared error (Figure \ref{fig:figure-1}D, dashed colored lines), but does
not entirely eliminate the error in the $2Ns \approx 1$ domain. Overall, while
we find evidence that local-rescaling slightly reduces error in the $2Ns
\approx 1$ domain, we do not use it during in our genome-wide inference method.
This is because to avoid circularity and possible overfitting, our method would
need to solve the equilibria equations for all segments simultaneously;
unfortunately, this is currently computational unfeasible.

@@

However, across all of our annotation models, new deleterious mutations in
conserved feature classes (CADD tracks, CDS, and PhastCons regions) were
consistently estimated to have strongly deleterious effects (Figure
\ref{fig:figure-4}A), consistent with previous work
\parencite{McVicker2009-ax,Murphy2022-sj}. The DFE estimates for CADD and
PhastCons regions consistently places $\ge 75\%$ of mass on the largest
selection coefficient, $s=10^{-2}$. The CADD 6\% DFE estimates imply an average
selection coefficient of $\bar{s} = 0.65\%$ for CEU, $\bar{s} = 0.57\%$ for
CHB, and $\bar{s} = 0.79\%$ for YRI. Similarly, the PhastCons Priority model
implies average selection coefficient estimates of $\bar{s} = 0.63\%$, $\bar{s}
= 0.58\%$, and $\bar{s}= 0.77\%$ for CEU, CHB, and YRI respectively for
PhastCons regions. Our DFE estimates for the Feature Priority model are
qualitatively consistent with the U-shaped DFEs found for amino acids through
Poisson Random Field method \parencite{Boyko2008-tj}.

Following previous work, our method used a grid of selection coefficients up to
$s=10^{-2}$. However, we also experimented with a strong selection grid that
includes $s=10^{-1}$. We find that models fit with the strong selection grid
had $R_\text{LOCO}^2$ that were about one percentage point higher, which is
suggestive of stronger selection than has previously been estimated using
constrained grids (see Supplementary Materials Table
\ref{supp:tbl-r2-strongsel}). For this strong selection grid, we estimate
average selection coefficients for the CADD 6\% model of $\bar{s} = 4.3\%$ for
CEU, $3.2\%$ for CHB, and $4.4\%$ for YRI. Thus, across different selection
coefficient grids, there is considerable uncertainty in the average strength of
selection against deleterious mutations in conserved regions.

However, across the strong selection grid runs, we find indications of model
non-identifiability. First, estimates of the DFE with the strong grid are
strangely bimodal (Supplementary Figure \ref{suppfig:strong-sel-dfe}). For
example, under the CADD 6\% strong selection grid model, new mutations are
estimated to have a selection coefficients of $s = 10^{-1}$ with 43\% chance,
$s=10^{-2}$ with 7.8\% chance, and $s=10^{-3}$ with 41\% chance. We propose
that one mechanism for this non-identifiability is that mutations extremely
harmful to fitness are difficult to distinguish from a smaller effective
population. Intuitively, as $s$ increases, the average persistence time (in
generations) of the mutation in the population goes below one generation, i.e.
death of the carrier of the \emph{de novo} mutation. In this hypothetical
limit, each occurrence of a very harmful mutation nearly acts like an
instantaneous death that in effect is just a smaller population size. One way
to test this hypothesis is to look to see if there is a systematic positive
relationship across models between average selection coefficient and $\pi_0$,
which is includes the drift-effective population size $N_e$.

We find this is the case for all of our CADD 6\% models. Across every
population, average selection was about $7.1$ times larger using the default
grid, and $\pi_0$ was $5.6\%$ larger (see Supplementary Materials Figure
\ref{suppfig:sel-ident}). There was no similar consistent change in mutation
rate estimates among populations. In the CADD 6\% model, genome-wide average
reduction factor $\bar{B}$ was $\approx 6.1\%$ lower in the default versus
constrained grid. Overall, this suggests that the linked selection signal alone
cannot differentiate very strong selection from a slightly smaller
drift-effective population size.

@@

Although the Feature Priority model fits the data less well than alternative
models, its features are more interpretable. We find that our B' method
estimates a bimodal DFE for coding sequences for the Feature Priority model,
with a large mass placed on $10^{-3} \le s \le 10^{-2}$ and another on the
neutral class $s=10^{-8}$. This is expected, given that the synonymous and
non-synonymous sites that constitute coding sequences are under vastly
different levels of constraint. Moreover, features expected to be only weakly
constrained such as introns and UTRs have the bulk of DFE mass on the neutral
class, with a small but significant amount of mass ($\approx 3\%$) placed on
$s=10^{-2}$. As expected, the DFE for PhastCons regions suggests it is under
strong selective constraint; however we note that block jackknife based
uncertainty estimates suggest the model is uncertain whether there is some mass
on the neutral class. Finally, we highlight one result from our PhastCons
Priority annotation model (Figure \ref{fig:figure-3}A bottom row): the DFE
estimate for coding sequences excluding PhastCons regions is estimated as
neutral. This too is expected; the selection signal in coding regions is
absorbed by the PhastCons feature, leaving only conditionally neutral sites.

@@

Early genome-wide implementations of the classic BGS model fit well, but lead
to unusually high estimates of the mutation rate \parencite{McVicker2009-ax}.
This lead to the hypothesis that these models could be absorbing the signal of
positive selection \parencite{Enard2014-kz}, though other work has found a
limited role for hitchhiking
\parencite{Pickrell2009-tt,Hernandez2011-gs,Murphy2022}. While our simulation
results suggest estimates of the mutation rate from linked selection models are
biased, we still our estimates were in rough accordance with pedigree-based
estimates \parencite{Kong2013-fc,Tian2019-so}.

We find across all populations, our mutation rate estimates from CADD-based
models are roughly consistent with pedigree-based estimates (Figure
\ref{fig:figure-4}A), consistent with recent work \parencite{Murphy2022-sj}.
Our full-track CADD 6\% model estimates the mutation rate as $\widehat{\mu} =
1.56 \times 10^{-8}$ for YRI, $1.63 \times 10^{-8}$ for CEU, and $1.60 \times
10^{-8}$ for CHB reference samples (Supplementary Table \ref{supp:tbl-r2}). As
expected, the sparse-track CADD model mutation rate estimates are identical
between the B' and classic BGS methods (Figure \ref{fig:figure-4}A top row).

However, mutation rate estimates for feature-based annotation models are less
concordant with pedigree-based estimates. First, mutation rate estimates under
our implementation of classic BGS theory are an order of magnitude below the
expected range (Figure \ref{fig:figure-4}A top row). We observe similar
behavior when we use the classic BGS model to fit full-coverage annotation
models (Figure \ref{fig:figure-4}A bottom row). This behavior is consistent
with classic BGS theory being unable to fit the DFE to features under weak
constraint (e.g. introns, UTRs, and the ``other" feature), and most compensate 
by estimating an unusually low mutation rate.

Second, we noticed that across all populations and sparse/full tracks, the CADD
6\% model consistently led to slightly higher mutation rates than the CADD 8\%
model (Figure \ref{fig:figure-4}A bottom row; Supplementary Table
\ref{supp:tbl-r2}). This pattern was observed in Murphy et al. too (2023;
Appendix 1, Figure 16). This behavior suggests a non-identifiability issue
between a slightly higher per-basepair mutation rate and annotation tracks that
contain more conserved sequence. This is expected from theory, since both
classic BGS and SC16 models only depend on mutation rate through the compound
parameter $\mu L$, where $L$ is the length of the conserved segment. Even
though our method is much more robust to including non-conserved regions like
introns, we still observe this non-identifiability issue.

Finally, we note that mutation rate estimates in the Feature Priority model are
unreasonably high ($\widehat{\mu} \approx 3 \times 10^{-8}$), reminiscent of
the high mutation rate estimates found under McVicker et al.'s
(\citeyear{McVicker2009-ax}) model. While both our and Murphy et al.'s CADD and
PhastCons-based models alleviate this issue, it is worth discussing why this
could occur. We gain some insight from comparing the estimated mutation rates
between our Feature and PhastCons Priority annotation models, which both
contain the exact same number of basepairs, but these are assigned to different
feature classes based on the priority of overlapping features. That one of
these models is our best-fitting model, and the other our worst indicates the
sensitivity these models have when features classes have heterogeneous DFEs.
CADD-based models fit better in part due to their fine-scale resolution of
selective effects across the genome. While ideally we would fit a CADD model
with different features corresponding to the different percentiles of
pathogenicity, these features are on the basepair scale and thus too
memory-intensive for our method to currently accommodate.

@@

To compare our models' predicted substitution rates against observed
divergence, we estimated sequence divergence on the human lineage using a
multiple alignment of five primates for each feature in our feature-based
models (Methods XXX). Since our simulations show that mutation rate estimates
can be biased, we predicted substitution rates under a fixed mutation rate of
$\mu = 1.5 \times 10^{-8}$. Fixing the mutation rate also allows us to more
easily compare the predictions across our feature-based models. Unfortunately,
a careful comparison between our predictions and observed divergence rates is
hindered by considerable uncertainty in the rate of the molecular clock,
generation times, and the human-chimpanzee divergence time. We assume a
generation time of 29 years, and calculate the sequence divergence implied by
our predicted substitution rates over a range of divergence times, from 6 Mya
to 12 Mya \parencite{Moorjani2016-tb,Nachman2000-te,Yi2002-pw,Steiper2006-xx}.

We find that our predicted substitution rates are qualitatively consistent with
the observed divergence along the human lineage for all features except
PhastCons (Figure \ref{fig:figure-7}).

As expected, the predicted substitution
rate in features unlikely to be under selective constraint is very close to the
mutation rate. Throughout, we report our substitution rates as a percent
relative to the total mutation rate, $\mu$ (here fixed to $1.5 \times
10^{-8}$). In our PhastCons Priority model, introns and UTRs are estimated to
have have substitution rates that are $93.6\%$ of the mutation rate, and the
``other" feature is estimated at $97.8\%$.  When we use the constrained grid,
these change to $97.85\%$ and $99.98\%$ respectively. In our PhyloFit
estimates, the divergence rate in introns is $92.44\%$, 5' UTRs $75.5\%$, and
3' UTRs $72.65\%$. By contrast, our predicted substitution rates for PhastCons
regions are much lower: $0.48\%$ for the default grid, and $1.65\%$ for the
constrained grid. For comparison, the PhyloFit estimates for PhastCons regions
is much higher, at $51.23\%$. Since this would include beneficial substitutions
that would not be predicted by our model, this suggests a very high proportion
of beneficial substitutions XXX check

% This is likely due to the constrained grid placing more of the DFE mass on
% weaker selection coefficients, which in turn have a higher chance of drifting
% up to fixation.

@@

\subsection*{Substitution Rate and Total Genic Fitness Variance Predictions}

a set of corresponding substitution rate predictions for each segment under
mutation-selection-draft process. There are two sources of variation in these
predictions in this model. First, there is the variation across the
pre-specified functional categories, e.g. CDS, UTRs, and PhastCons. Second,
there is region-specific variation

The SC16 theory models linked selection in quantitative genetics terms and
relies on simultaneously predicting the deleterious substitution rate.

>>>>

The discrepancy between our theoretic B' maps and chromosome-wide simulations
that leads to the 5\% error occurs entirely around the drift-barrier domain of
$2Ns \approx 1$ (Figure \ref{fig:figure-1}D). The higher error in this domain
occurs across all mutation rates. We hypothesized this error may be due to
selective interference between segments that is not taken into account when we
numerically solve Equations \eqref{eq:main_eqns} for each segment
independently. In particular, we use a fixed drift-effective $N=1000$
corresponding to the number of diploids in the simulations rather than $B(x) N$
at position $x$. To test this, we implemented a ``locally-rescaled" version of
the B' maps, which numerically solves Equations \eqref{eq:main_eqns} for each
segment at position $x$ taking into account the reduction experienced due to
selection at other segments by setting $B'(x)N$ as the population size. We find
the locally-rescaled B' maps reduce the relative error from 5\% to 0.4\% and
mean squared error (Figure \ref{fig:figure-1}D, dashed colored lines), but does
not entirely eliminate the error in the $2Ns \approx 1$ domain. Overall, while
we find evidence that local-rescaling slightly reduces error in the $2Ns
\approx 1$ domain, we do not use it during in our genome-wide inference method.
This is because to avoid circularity and possible overfitting, our method would
need to solve the equilibria equations for all segments simultaneously;
unfortunately, this is currently computational unfeasible.

@@

However, across all of our annotation models, new deleterious mutations in
conserved feature classes (CADD tracks, CDS, and PhastCons regions) were
consistently estimated to have strongly deleterious effects (Figure
\ref{fig:figure-4}A), consistent with previous work
\parencite{McVicker2009-ax,Murphy2022-sj}. The DFE estimates for CADD and
PhastCons regions consistently places $\ge 75\%$ of mass on the largest
selection coefficient, $s=10^{-2}$. The CADD 6\% DFE estimates imply an average
selection coefficient of $\bar{s} = 0.65\%$ for CEU, $\bar{s} = 0.57\%$ for
CHB, and $\bar{s} = 0.79\%$ for YRI. Similarly, the PhastCons Priority model
implies average selection coefficient estimates of $\bar{s} = 0.63\%$, $\bar{s}
= 0.59\%$, and $\bar{s}= 0.77\%$ for CEU, CHB, and YRI respectively for
PhastCons regions. Our DFE estimates for the Feature Priority model are
qualitatively consistent with the U-shaped DFEs found for amino acids through
Poisson Random Field method \parencite{Boyko2008-tj}.

Following previous work, our method used a grid of selection coefficients up to
$s=10^{-2}$. However, we also experimented with a strong selection grid that
includes $s=10^{-1}$. We find that models fit with the strong selection grid
had $R_\text{LOCO}^2$ that were about one percentage point higher, which is
suggestive of stronger selection than has previously been estimated using
constrained grids (see Supplementary Materials Table
\ref{supp:tbl-r2-strongsel}). For this strong selection grid, we estimate
average selection coefficients for the CADD 6\% model of $\bar{s} = 4.2\%$ for
CEU, $3.2\%$ for CHB, and $4.4\%$ for YRI. Thus, there is quite a degree of
sensitivity to DFEs and average selection estimates based on the selection
coefficient grid.

However, across the strong selection grid runs, we find indications of model
non-identifiability. First, estimates of the DFE with the strong grid are
strangely bimodal (Supplementary Figure \ref{suppfig:strong-sel-dfe}). For
example, under the CADD 6\% strong selection grid model, new mutations are
estimated to have a selection coefficients of $s = 10^{-1}$ with 43\% chance,
$s=10^{-2}$ with 7.8\% chance, and $s=10^{-3}$ with 41\% chance in the YRI
samples. We propose that one mechanism for this non-identifiability is that
very deleterious mutations lead to larger whole-genome reductions, which are
difficult to distinguish from a smaller drift effective population size (i.e.
the $\pi_0$ parameter). One way to test this hypothesis is to look to see if
there is a systematic positive relationship across models between average
selection coefficient and $\pi_0$, which is includes the drift-effective
population size $N_e$.

We find this is the case for all of our CADD 6\% models. Across all
populations, average selection was about $7.1$ times larger using the strong
selection grid, and $\pi_0$ was $5.6\%$ higher (see Supplementary Materials
Figure \ref{suppfig:sel-ident}). There was no similar consistent change in
mutation rate estimates among populations. In the CADD 6\% model, genome-wide
average reduction factor $\bar{B}$ was $\approx 6.1\%$ lower in the default
versus constrained grid. Overall, this suggests that the linked selection
signal alone cannot differentiate very strong selection from a slightly smaller
drift-effective population size.

@@

Although the Feature Priority model fits the data less well than alternative
models, its DFE estimates are are more interpretable. We find that our B'
method estimates a bimodal DFE for coding sequences for the Feature Priority
model, with a large mass placed on $10^{-3} \le s \le 10^{-2}$ and another on
the neutral class $s=10^{-8}$. This is expected, given that the synonymous and
non-synonymous sites that constitute coding sequences are under vastly
different levels of constraint. Moreover, features expected to be only weakly
constrained such as introns and UTRs have the bulk of DFE mass on the neutral
class, with a small but significant amount of mass ($\approx 3\%$) placed on
$s=10^{-2}$. As expected, the DFE for PhastCons regions (which in this model
correspond to highly-conserved non-coding elements) suggests it is under strong
selective constraint; however we note that block jackknife based uncertainty
estimates suggest the model is uncertain whether there is some mass on the
neutral class. Finally, we highlight one result from our PhastCons Priority
annotation model (Figure \ref{fig:figure-3}A bottom row): the DFE estimate for
coding sequences excluding PhastCons regions is estimated as neutral. This too
is expected; the selection signal in coding regions is absorbed by the
PhastCons feature, leaving only conditionally neutral sites.

@@

Prior work on genome-wide inference using the classic BGS model fit the linked
selection signal well, but led to unusually high estimates of the mutation rate
\parencite{McVicker2009-ax}. This lead to the hypothesis that these models
could be absorbing the signal of positive selection \parencite{Enard2014-kz},
though other work has found a limited role for hitchhiking
\parencite{Pickrell2009-tt,Hernandez2011-gs,Murphy2022}. While our simulation
results suggest estimates of the mutation rate from linked selection models are
biased, we still check for rough agreement with pedigree-based estimates
\parencite{Kong2013-fc,Tian2019-so}.

We find across all populations, our mutation rate estimates from CADD-based
models are roughly consistent with pedigree-based estimates (Figure
\ref{fig:figure-4}A), consistent with recent work \parencite{Murphy2022-sj}.
Our full-track CADD 6\% model estimates the mutation rate as $\widehat{\mu} =
1.56 \times 10^{-8}$ for YRI, $1.64 \times 10^{-8}$ for CEU, and $1.60 \times
10^{-8}$ for CHB reference samples (Supplementary Table \ref{supp:tbl-r2}). As
expected, the sparse-track CADD model mutation rate estimates are nearly
identical between the B' and classic BGS methods (Figure \ref{fig:figure-4}A
top row).

However, mutation rate estimates for feature-based annotation models are
implausibly higher than pedigree-based estimates. First, mutation rate
estimates under our implementation of classic BGS theory are an order of
magnitude below the expected range (Figure \ref{fig:figure-4}A top row). We
observe similar behavior when we use the classic BGS model to fit full-coverage
annotation models (Figure \ref{fig:figure-4}A bottom row). This behavior is
consistent with classic BGS theory being unable to fit the DFE to features
under weak constraint (e.g. introns, UTRs, and the ``other" feature), and most
compensate by estimating an unusually low mutation rate.

Second, we noticed that across all populations and sparse/full tracks, the CADD
6\% model consistently led to slightly higher mutation rates than the CADD 8\%
model (Figure \ref{fig:figure-4}A bottom row; Supplementary Table
\ref{supp:tbl-r2}). This same pattern was observed in Murphy et al. too (2023;
Appendix 1, Figure 16). This behavior suggests a non-identifiability issue
between a slightly higher per-basepair mutation rate and annotation tracks that
contain more conserved sequence. This is expected from theory, since both
classic BGS and SC16 models only depend on mutation rate through the compound
parameter $\mu L$, where $L$ is the length of the conserved segment. Even
though our method is much more robust to inclusion of non-conserved regions
like introns, we still observe this non-identifiability issue.

Finally, we note that mutation rate estimates in the Feature Priority model are
unreasonably high ($\widehat{\mu} \approx 3 \times 10^{-8}$), reminiscent of
the high mutation rate estimates found under McVicker et al.'s
(\citeyear{McVicker2009-ax}) model. While both our and Murphy et al.'s CADD and
PhastCons-based models alleviate this issue, it is worth discussing why this
could occur. We gain some insight from comparing the estimated mutation rates
between our Feature and PhastCons Priority annotation models, which both
contain the exact same number of feature basepairs, but these are assigned to
different feature classes based on the priority of overlapping features. That
one of these models is our best-fitting model, and the other our worst
indicates the sensitivity these models have when features classes have
heterogeneous DFEs. CADD-based models fit better in part due to their
fine-scale resolution of selective effects across the genome. While ideally we
would fit a CADD model with different features corresponding to the different
percentiles of pathogenicity, these features are on the basepair scale and thus
too memory-intensive for our method to currently accommodate.

@@

To compare our models' predicted substitution rates against observed
divergence, we estimated sequence divergence on the human lineage using a
multiple alignment of five primates for each feature in our feature-based
models (Methods \ref{sec:methods-sub}). We compared these to the predicted
substitution rates per feature, averaging over all segments in the genome.
Since our simulations show that mutation rate estimates can be biased, we
predicted substitution rates under a fixed mutation rate of $\mu = 1.5 \times
10^{-8}$. Fixing the mutation rate also allows us to more easily compare the
predictions across our feature-based models. Unfortunately, a careful
comparison between our predictions and observed divergence rates is hindered by
considerable uncertainty in the rate of the molecular clock, generation times,
and the human-chimpanzee divergence time. We assume a generation time of 29
years, and calculate the sequence divergence implied by our predicted
substitution rates over a range of divergence times, from 6 Mya to 12 Mya
\parencite{Moorjani2016-tb,Nachman2000-te,Yi2002-pw,Steiper2006-xx}.

We find that our predicted substitution rates are qualitatively consistent with
the observed divergence along the human lineage for all features except
PhastCons (Figure \ref{fig:figure-7}). As expected, the predicted substitution
rates in features unlikely to be under selective constraint (introns and UTRs,
and the ``other" feature) are very close to the mutation rate. Throughout, we
report our substitution rates as a percent relative to the total mutation rate,
$\mu$ (here fixed to $1.5 \times 10^{-8}$). In our Feature Priority model,
coding sequences are predicted to have a substitution rate of 41.20\% of the
mutation rate, introns and UTRs 94.71\%, coding sequences 0\%, and the ``other"
feature 99.98\%. For comparison, the substitution rates along the human lineage
(as a proportion to the substitution rate in putatively neutral regions) are
74.15\% in UTRs, 92.44\% in introns, 50.96\% in coding sequences, and 49.56\%
in PhastCons regions. The large discrepancy between predicted and observed
PhastCons substitution rates is driven by our DFE estimates suggesting that the
bulk of mass is on selection coefficients greater than $10^{-3}$, which have no
chance of fixation in a population of $N_e \approx 10,000$. We note that our
DFE estimates are qualitatively similar to those inferred using the classic BGS
model; thus this DFE-substitution rate contradiction is not unique to our
estimates, and could indicate a potential model misspecfication problem.

XXX HERE

@@

\subsection*{Substitution Rate Estimation and Prediction}
\label{sec:methods-sub}

We estimated the divergence along the human branch using PhyloFit
\parencite{Siepel2004-ql} run on the Multiz alignments
\parencite{Blanchette2004-bb} of humans and four other primates (\emph{Pongo
abelii}, \emph{Pan troglodytes}, \emph{Pan paniscus}, \emph{Gorilla gorilla}).
PhyloFit was run using the HKY85 substitution model per-feature; estimates from
alternate substitution models yielded equivalent results. Further details about
this process can be be found in the GitHub repository
(\url{https://github.com/vsbuffalo/bprime}).
